pSEOBNRv4HM

### 2020-10-20 papers/intro_paper/intro_paper.tex
updated paper draft

Observation of gravitational waves have had significant astrophysical and cosmological implications~\citep{LSC_2016astroph,gw170817_mma,gw170817_joint,gw170817_hubble}. It has also allowed us to make statements in fundamental physics. Specifically, LIGO-Virgo's observations have allowed us to test predictions of Einstein's theory of General Relativity~\citep[GR]{}, in previously unexplored regimes of highly relativistic, strong-field regimes of gravity~\citep{LSC_2016grtests,GW170817_TGR,gwtc1_tgr}. In GR, a CBC involving two black holes (a binary black hole/BBH system) is described in three distinct phases: an early \textit{inspiral}, where the two compact objects spiral in and \textit{plunge} due to a back-reaction of gravitational-wave-emission, a \textit{merger} event marked by the formation of an apparent horizon~\citep{NRpaper}, and a late-time \textit{ringdown}, where the newly formed remnant object settles down to a stable Kerr state through the emission of an exponentially damped quasi-normal-mode (QNM) spectrum of gravitational radiation~\citep{vishu,earlyqnmpapers}.  We have performed tests of gravitational wave generation and source dynamics, where we place bounds on parametrised deviations in the Post-Newtonian coefficients describing the early inspiral~\citep{earlydevelopmentpapers}, and phenomenological coefficients describing the intermediate (plunge) and merger regimes of coalescence~\citep{TIGERmethodspapers}, tests of gravitational wave propagation, which assume a generalised dispersion relation and place upper bounds on the Compton wavelength and consequently, the mass of the graviton~\citep{gw170104,samajdar2017projected} and tests of the polarisation of gravitational radiation using a multi-gravitational-wave-detector network~\citep{gw170814,isi2017probing}. We have also checked for consistency between different portions of the signal using estimates of final mass and spin~\citep{Ghosh:2016xx,Ghosh:2017gfp,LSC_2016grtests}, and consistency of the residuals with interferometric noise~\citep{Ghonge:2020suv,gwtc1_tgr}. None of these tests report any departure from the predictions of GR.

@@

Tests of black hole ringdown and nature of the remnant object is an active field of research at the moment. The no-hair conjecture in GR~\citep{} states that an (electrically neutral) astrophysical black hole is completely described by two observables: mass and spin angular momentum. One consequence of the no-hair conjecture is that the (complex) QNM frequencies of gravitational radiation emitted by a perturbed isolated black hole is uniquely determined by its mass and spin angular momentum. Hence a test of the no-hair conjecture would involve checking for consistency between estimates of mass and spin of the remnant object across multiple QNM frequencies. An inconsistency would either indicate a non-black hole nature of the remnant object, or an incompleteness of GR as a theory of gravity. The consistency between the post-merger signal and the least dampled QNM was first demonstrated in ~\citep{LSC_2016grtests}, and later extended to include overtones in~\citep{Giesler:2019uxc,Isi:2019aib,Bhagwat:2019dtm,Forteza:2020hbw}. Consistency of the late-time signal with a single QNM is a test of the ringdown of a BBH coalescence, but not necessarily a test of the no-hair conjecture, which requires the measurement of (atleast) two QNMs (black hole spectroscopy), and checking for consistency between them. Recent work in that direction include~\citep{Carullo:2018gah,Carullo:2019flw,Bhagwat:2019bwv}. The nature of the remnant object has also been explored through tests of black hole thermodynamics, like the Hawking's area theorem~\citep{Cabero:2017avf} or through search for echos in the post-merger signal~\citep{Nielsen:2018lkf,Tsang:2019zra,Lo:2018sep,Abedi:2018npz,Abedi:2020sgg,Testa:2018bzd}. None of these tests have found evidence for non-black hole nature of the remnant object (as described in GR) in LIGO-Virgo BBH observations.

@@

Unlike methods that focus on the post-merger signal, it describes a framework in which a complete inspiral-merger-ringdown (IMR) waveform model is used to measure the complex QNM frequencies. \abhi{It is a spinning multipolar effective-one-body waveform calibrated to numerical relativity simulations (SEOBNRv4HM~\citep{Cotesta:2018fcv}})This allows access to the full SNR of the signal, reducing measurement uncertainties. Moreover, the definition of the ringdown start time is built into the model and does not need to be either left as an additional free parameter or fixed using alternate definitions. While \paperone presented the method in the context of a non-spinning waveform model, we extend the analysis to the case with black holes spins in the current paper. All astrophysical black holes are expected to be spinning, and ignoring effects of spin have been shown to introduce systematic biases in the measurement of the source properties.
\par
The rest of this paper is organized as follows. Section~\ref{sec:model} describes our parametrised IMR waveform model. In section~\ref{sec:method}, we define our testing GR framework in which we use that model to measure the complex QNM frequencies in a Bayesian formalism. Then in section~\ref{sec:results}, we demonstrate our method on real gravitational wave events, as well as simulated signals. Finally we provide a summary of our results and discuss future developments in section  The rest of this paper is organized as follows. Section~\ref{sec:model} describes our waveform model, which is then used to perform our test of black hole ringdown using Bayesian framwork, outlined in section~\ref{sec:discussion}.

@@

Throughout our analysis, we assume the noise in the detector to be well-approximated by the Gaussian random process with a fixed power spectral density. This allows us to define a Gaussian likelihood function $\mathcal{L}(d | \blambda, H, I)$, which we stochastically sample over using either a Markhov-Chain-Monte-Carlo based algorithm implemented in the LIGO Algorithms Library.

\section{Results}\label{sec:results}
\subsection{Priors}

@@

\subsection{Results on GW150914-like injection}

\begin{figure*}
	\includegraphics[width=\textwidth]{figures/GW150914_swinj_GR_vs_modGR.png}
	\caption{final figure will contain a third set of contours for modGR recovered assuming GR to show bias. That run is ongoing.}
\end{figure*}

\abhi{The GW150914-like injections we have:}
\begin{itemize}
\item{an SEOBNRv4HM software injection with the 22 mode frequencies estimated}
\item{a non-spinning NR injection with parameters close to the real event but higher SNR for which 22 mode frequencies are estimated}
\item{a non-spinning highly-asymmetric (q=6) NR injection with total mass similar to GW150914 and high SNR for which 22 and 33 mode frequencies are estimated}
\end{itemize}

\subsection{Results on modGR injection}

\abhi{As part of review we had developed the infrastructure to make non-GR waveforms using pSEOBNRv4HM, and running LALInference on it. We had specifically chosen an injection with domega220 = dtau220  = 0.5 in zero-noise, and successfully recovered those frequencies thus demonstrating the robustness of the method in detecting deviations from GR.}

@@

\begin{figure}
	\includegraphics[width=0.4\textwidth]{figures/GW150914.png}
	\caption{include contours from pyRING}
\end{figure}

\abhi{Results obtained as part of the review on GW150914 and other O1-O2 events which are of relevance}
\abhi{A plot showing the fractional quantities}
\abhi{A table with absolute quantites similar to GWTC-2 TGR}

@@

\bottomrule
\end{tabular}

@@

\abhi{A small summary of the noise systematics study that was done for S190521g to show that the deviations seen were consistent with systematics due to noise. The study include 2 separate sections: in Gaussian noise, and in real LIGO-Virgo noise immediately adjacent to the actual event.}

%\begin{figure}[H]
%	\includegraphics[width=0.4\textwidth]{figures/S190521g_noiseinjections.png}
%	\caption{a single figure for both Gaussian and Real noise injections, color coded, to save space}
%\end{figure}

We injected NRSur7dq4 signal using the MaxL parameters from EXP30, in 2.5 hours of data surrounding the event at: t0-1.5 hours, t0-1., t0-0.5, t0+0.5, t0+1. For 3 of the 5 noise realisations, corresponding to t0-1 hour, t0+0.5 hours, and t0+1.0 hour we recover a damping time similar to the actual event, whereas for a third case, t0-0.5 hours, we obtain a damping time consistent at the 90\% credible interval. L1-SNR goes down by more than 3 in some runs (maximum L1-SNR 12, minimum 9), indicative of how the noise variability is strongly affecting the recovered parameters. For some injections the mass ratio is biased, although final spin (most sensitive to mass ratio) is not affected much.  domega varies quite a lot among runs, instead dtau seems well behaved around zero. Effective frequency is a stable parameter (always well measured), while effective tau is much more sensitive to the noise configuration.

>>>>

Observation of gravitational waves have had significant astrophysical and cosmological implications~\citep{LSC_2016astroph,gw170817_mma,gw170817_joint,gw170817_hubble}. It has also allowed us to make statements in fundamental physics. Specifically, LIGO-Virgo's observations have allowed us to test predictions of Einstein's theory of General Relativity~\citep[GR]{}, in previously unexplored regimes of highly relativistic, strong-field regimes of gravity~\citep{LSC_2016grtests,GW170817_TGR,gwtc1_tgr}. In GR, a CBC involving two black holes (a binary black hole (BBH) system) is described in three distinct phases: an early \textit{inspiral}, where the two compact objects spiral in and \textit{plunge} due to a back-reaction of gravitational-wave-emission, a \textit{merger} marked by the formation of an apparent horizon~\citep{NRpaper}, and a late-time \textit{ringdown}, where the newly formed remnant object settles down to a stable Kerr state through the emission of an exponentially damped quasi-normal-mode (QNM) spectrum of gravitational radiation~\citep{vishu,earlyqnmpapers}.  We have performed tests of gravitational wave generation and source dynamics, where we place bounds on parametrised deviations in the Post-Newtonian coefficients describing the early inspiral~\citep{earlydevelopmentpapers}, and phenomenological coefficients describing the intermediate (plunge) and merger regimes of coalescence~\citep{TIGERmethodspapers}, tests of gravitational wave propagation, which assume a generalised dispersion relation and place upper bounds on the Compton wavelength and consequently, the mass of the graviton~\citep{gw170104,samajdar2017projected} and tests of the polarisation of gravitational radiation using a multi-gravitational-wave-detector network~\citep{gw170814,isi2017probing}. We have also checked for consistency between different portions of the signal using estimates of final mass and spin~\citep{Ghosh:2016xx,Ghosh:2017gfp,LSC_2016grtests}, and consistency of the residuals with interferometric noise~\citep{Ghonge:2020suv,gwtc1_tgr}. None of these tests report any departure from the predictions of GR. \abhi{Perhaps reframe this section in the contect of the O3a TGR companion paper}

@@

Tests of black hole ringdown and nature of the remnant object is an active field of research at the moment. The no-hair conjecture in GR~\citep{} states that an (electrically neutral) astrophysical black hole is completely described by two observables: mass and spin angular momentum. One consequence of the no-hair conjecture is that the (complex) QNM frequencies of gravitational radiation emitted by a perturbed isolated black hole is uniquely determined by its mass and spin angular momentum. Hence a test of the no-hair conjecture would involve checking for consistency between estimates of mass and spin of the remnant object across multiple QNM frequencies. An inconsistency would either indicate a non-black hole nature of the remnant object, or an incompleteness of GR as the underlying theory of gravity. The consistency between the post-merger signal and the least dampled QNM was first demonstrated in ~\citep{LSC_2016grtests}, and later extended to include overtones in~\citep{Giesler:2019uxc,Isi:2019aib,Bhagwat:2019dtm,Forteza:2020hbw}. Consistency of the late-time signal with a single QNM is a test of the ringdown of a BBH coalescence, but not necessarily a test of the no-hair conjecture, which requires the measurement of (atleast) two QNMs (black hole spectroscopy), and checking for consistency between them. Recent work in that direction include~\citep{Carullo:2018gah,Carullo:2019flw,Bhagwat:2019bwv}. The nature of the remnant object has also been explored through tests of black hole thermodynamics, like the Hawking's area theorem~\citep{Cabero:2017avf} or through search for echos in the post-merger signal~\citep{Nielsen:2018lkf,Tsang:2019zra,Lo:2018sep,Abedi:2018npz,Abedi:2020sgg,Testa:2018bzd}. None of these tests have found evidence for non-black hole nature of the remnant object (as described in GR) in LIGO-Virgo BBH observations.

@@

Unlike methods that focus on the post-merger signal, it describes a framework in which a complete inspiral-merger-ringdown (IMR) waveform model is used to measure the complex QNM frequencies. \abhi{It is a spinning multipolar effective-one-body waveform calibrated to numerical relativity simulations (SEOBNRv4HM~\citep{Cotesta:2018fcv}})This allows access to the full SNR of the signal, reducing measurement uncertainties. Moreover, the definition of the ringdown start time is built into the model and does not need to be either left as an additional free parameter or fixed using alternate definitions. While \paperone presented the method in the context of a non-spinning waveform model, we extend the analysis to the case with black holes spins in the current paper. All astrophysical black holes are expected to be spinning, and ignoring effects of spin have been shown to introduce systematic biases in the measurement of the source properties.
\par
The rest of this paper is organized as follows. Section~\ref{sec:model} describes our parametrised IMR waveform model. In section~\ref{sec:method}, we define our testing GR framework in which we use that model to measure the complex QNM frequencies in a Bayesian formalism. Then in section~\ref{sec:results}, we demonstrate our method on real gravitational wave events, as well as simulated signals. Finally we provide a summary of our results and discuss future developments in section~\ref{sec:discussion}.

@@

An example of a posterior distribution on $\bzeta$Throughout our analysis, we assume the noise in the detector to be well-approximated by the Gaussian random process with a fixed power spectral density. This allows us to define a Gaussian likelihood function $\mathcal{L}(d | \blambda, H, I)$, which we stochastically sample over using either a Markhov-Chain-Monte-Carlo based algorithm implemented in the LIGO Algorithms Library.

\section{Results}\label{sec:results}

\iffalse
\subsection{Priors}

@@

\fi

\subsection{Simulations using GR signals}\label{ssec:gr_signal}

We initially demonstrate our method using simulated GR signals, with parameters similar to the two gravitational wave events, GW150914 CITE and GW190521 CITE. These two systems are representative data points for the kind of systems this test is most suitable for: high-mass events which are loud enough that the signal-to-noise ratio pre- and post-merger return reliable parameter estimation results. To avoid possible systematic biases in our parameter estimation results due to error in modelling, we use the GR version of the same waveform, SEOBNRv4HM CITE (without allowing for deviations in the QNM parameters) to simulate our GW signal. And to avoid systematic biases due to noise, we use an averaged (zero-noise) realisation of the Advanced LIGO zero-detuned-high-power PSD CITE. We find, as expected, that the posterior distribution on the parameters describing fractional deviations in the frequency and damping time, $\{d\omega,d\tau\}$ are consistent with zero (the solid curves in left panel of Fig.\ref{fig:gr_simulation}). The difference in the size of the posterior is due to the difference in the SNRs of the two events ($\sim 25$ for GW150914 as compared to $\sim 13$ for GW190521). One can then convert these fractional quantities into absolute quantities using the relations:

\begin{eqnarray}
f_{220}^{\text{eff}} &:=& f_{220}^{\text{GR}} (1 + \delta f_{220}) \\
\tau _{220}^{\text{eff}} &:=& \tau _{220}^{\text{GR}} (1 + \delta \tau_{220})
\end{eqnarray}

and construct posterior distributions on these effective quantities, $(f_{220}^{\text{eff}}, \tau _{220}^{\text{eff}})$ (solid lines in middle and right panels of Fig.\ref{fig:gr_simulation}). The GR quantities $(f_{220}^{\text{GR}}, \tau _{220}^{\text{GR}})$ are constructed using numerical relativity predictions of the mass and spin of the remnant object, starting from estimates of the masses and spins of the initial binary \abhi{add description of the contruction of GR quantities with details of NR fit formula}. In each of these cases, recovered posterior are consistent with the GR predictions (as indicated by the plus signs).

\begin{figure*}[h]
	\includegraphics[width=0.3\textwidth]{figures/GW150914_GW190521_simulated_frac.png}\includegraphics[width=0.56\textwidth]{figures/GW150914_GW190521_simulated_abs.png}\label{fig:gr_simulation}
	\caption{final figure will contain a third set of contours for modGR recovered assuming GR to show bias. That run is ongoing.}
\end{figure*}

\iffalse
\abhi{The GW150914-like injections we have:}
\begin{itemize}
\item{an SEOBNRv4HM software injection with the 22 mode frequencies estimated}
\item{a non-spinning NR injection with parameters close to the real event but higher SNR for which 22 mode frequencies are estimated}
\item{a non-spinning highly-asymmetric (q=6) NR injection with total mass similar to GW150914 and high SNR for which 22 and 33 mode frequencies are estimated}
\end{itemize}
\fi

\subsection{Simulations using nonGR signals}

To demonstrate the robustness of the method in detecting possible deviations of the underlying GW signal from predictions of GR, we inject simulated GW signals which are identical to the corresponding GR prediction upto merger \abhi{be more specific}, and differ in their post-merger description. We again choose systems with initial-binary-parameters similar to GW150914 and GW190521, but we attach a phenomenological post-merger signal described by deviation parameters $d\omega =d\tau = 0.5$. In other words, we assume that the frequency and damping time of our nonGR signal is 1.5 times the corresponding GR prediction, although the pre-merger signal is identical to GR (Fig.\ref{fig:nongr_waveform}). We also avoid waveform and noise systematic biases by choosing a configuration identical to the simulations described in Sec.\ref{ssec:gr_signal}. The posterior probability distributions on $\{d\omega,d\tau\}$ or equivalently$(f_{220}^{\text{eff}}, \tau _{220}^{\text{eff}})$ (dashed lines in Fig.\ref{fig:gr_simulation})) are consistent with the corresponding the values of the injection parameters, indicated by plus signs \abhi{add markers in effective quantity plots}.

\begin{figure}
	\includegraphics[width=0.5\textwidth]{figures/modGR_waveforms.png}\label{fig:nongr_waveform}
	\caption{NonGR waveforms}
\end{figure}

\subsection{Simulations using NR signals}

\iffalse
\abhi{As part of review we had developed the infrastructure to make non-GR waveforms using pSEOBNRv4HM, and running LALInference on it. We had specifically chosen an injection with domega220 = dtau220  = 0.5 in zero-noise, and successfully recovered those frequencies thus demonstrating the robustness of the method in detecting deviations from GR.}
\fi

@@

The LIGO-Virgo Collaboration recently released their testing GR catalog containing results of this test for all events observed during the first half of O3 CITE, which passed a threshold for the total detector frame mass > XX and SNRs in the pre- and post-merger regions to be atleast YY. For reasons of computational expense the analyses of the relevant events during the LIGO-Virgo's first and second observing runs could not be included in the catalog, and are being presented here for the first time. We conduct analysis on the 5 relevant events that pass the above thresholds from the O1-O2 catalog CITE: GW150914, GW170104, GW170729, GW170814 and GW170823 \abhi{what about GW170809?}. Out of these, the analysis of GW170814 and GW170823 return uninformative posteriors. This is because of an insufficient SNR In the post-merger signal for these two events. For GW170814, this is because the total mass is slightly lower than the rest, thus coalescing in the region of the detector bandwidth dominated by shot-noise, while in the case of GW170823 the overall SNR of the entire signal is too low to reliably estimate $\{d\omega,d\tau\}$. For the three remaining signals, GW150914, GW170104 and GW170729, we provide the posterior distributions in $\{d\omega,d\tau\}$ in the top panel of Fig.\ref{fig:o1o2_events}. We also reconstruct the effective QNM parameters, $(f_{220}^{\text{eff}}, \tau _{220}^{\text{eff}})$ which are tabulated in Table \ref{tab:qnm_o1o2_results}.

Among all GW signals detected so far, GW150914 is unique in it's loudness, mass as well as the clarity of the signal, leading to the first, and to date, best attempt in measuring the QNM frequencies CITE. The bottom panel of Fig.\ref{fig:o1o2_events} compares the results we obtain for GW150914 with previously published LIGO-Virgo results using a damped-sinusoid method, as well as the measurement reported using a non-spinning model in \paperone. The damped-sinusoid model assumed specific cutoff times (indicated in the curves on the bottom panel of Fig.\ref{fig:o1o2_events}) after the time corresponding to the peak amplitude as predicted by the best-fitting GR template waveform. Increasing time, decreasing SNR, wider posteriors, less-precise measurement. Compared to that we see when we incorporate the complete IMR version of the test, we utilise more SNR leading to tighter constraints.

\begin{figure}
	\includegraphics[width=0.4\textwidth]{figures/O1O2_realevents.png}
	\includegraphics[width=0.4\textwidth]{figures/GW150914.png}
	\caption{O1-O2 analysis}\label{fig:o1o2_events}
\end{figure}

@@

\bottomrule
\end{tabular}\label{tab:qnm_o1o2_results}

@@

\iffalse
\abhi{A small summary of the noise systematics study that was done for S190521g to show that the deviations seen were consistent with systematics due to noise. The study include 2 separate sections: in Gaussian noise, and in real LIGO-Virgo noise immediately adjacent to the actual event.}
\fi

Tests of general relativity

\begin{figure}[H]
	\includegraphics[width=0.4\textwidth]{figures/S190521g_swinjs.png}
	\caption{a single figure for both Gaussian and Real noise injections, color coded, to save space}\label{fig:21g_systematics}
\end{figure}

We injected NRSur7dq4 signal using the MaxL parameters from EXP30, in 2.5 hours of data surrounding the event at: t0-1.5 hours, t0-1., t0-0.5, t0+0.5, t0+1. For 3 of the 5 noise realisations, corresponding to t0-1 hour, t0+0.5 hours, and t0+1.0 hour we recover a damping time similar to the actual event, whereas for a third case, t0-0.5 hours, we obtain a damping time consistent at the 90\% credible interval (bottom panel of FIg.\ref{fig:21g_systematics}). L1-SNR goes down by more than 3 in some runs (maximum L1-SNR 12, minimum 9), indicative of how the noise variability is strongly affecting the recovered parameters. For some injections the mass ratio is biased, although final spin (most sensitive to mass ratio) is not affected much.  domega varies quite a lot among runs, instead dtau seems well behaved around zero. Effective frequency is a stable parameter (always well measured), while effective tau is much more sensitive to the noise configuration.
